### 2020-05-13 docs/ICTT2019/RMethod_ictt26.tex
Upgrade AA by DAAREM (without the use of a merit function of interest); fix b.c. and transmission/reflection in the balance equation for the current in CPM1D - this includes correcting the documentation; implement the homogeneous cases by Sood PUa-x/PUb-x in both transport and diffusion+RM; implement the RM through escape probabilities; verify the CPM1D solver in the homogeneous cases; reorganize sources by adding a new module to manage material and geometry data.

%
% FIRST AUTHORS
%
\setlength{\baselineskip}{12pt}
\textbf{Daniele Tomatis}\\
DEN, Service d'{\'e}tudes des r\'eacteurs et de math\'ematiques appliqu\'ees (SERMA),\\
CEA, Universit\'e Paris-Saclay, F-91191 Gif-sur-Yvette, France\\
daniele.tomatis@cea.fr\\
\vspace{0.25in}
\textbf{Roy Gross and Erez Gilad}\\
The Unit of Nuclear Engineering\\
Ben-Gurion University of the Negev, Israel\\
roygros@post.bgu.ac.il, gilade@bgu.ac.il\\
\end{center}

@@

Particle conservation on the full domain requires $\sum_i \tilde{p}_{ij} + \tilde{e}^+_{I+1/2,j} = 1$, $\forall j$. At last, reciprocity implies
\begin{subequations}
\label{eq:reciprocity}
\begin{gather}
  \frac{\tilde{p}_{ij}}{V_i \sigma_i} = \frac{\tilde{p}_{ji}}{V_j \sigma_j} \quad \text{and} \label{eq:reciprocity1}\\
  \frac{\tilde{p}_{j, i+1/2}}{V_j \sigma_j} = \frac{\tilde{e}^+_{i+1/2, j} + \tilde{e}^-_{i+1/2, j}}{S_{i+1/2}}, % / 4 ?
  \label{eq:reciprocity2}
\end{gather}
\end{subequations}
for $1 \leq i,j, \leq I$. % and with the escape probability counting for both contributions from partial currents. These expressions
Eq. \ref{eq:reciprocity1} can also be used to verify the implementation of the escape probabilities. Collision probabilities with surface source emission from Eq. \ref{eq:reciprocity2} are needed with non-zero entering currents of neutrons at the boundary $x_{I+1/2}$. This probability refers to an entering angular flux that must be still divided by 4 in case the flux is isotropic, yielding a cosine current typical of white reflection.

@@

A generalized form for the boundary condition of the diffusion equation applying at the left side of the slab follows as $\Upsilon = - D_0 \phi_0 / (\Delta_0 / 2 + \zeta)$, where $\zeta$ is the extrapolation length in case of vacuum. Reflection can be reproduced by $\zeta \rightarrow \infty$, whereas the condition of zero-flux comes with $\zeta = 0$. The boundary $\delta J$ takes the simpler form $\delta J = - \delta D_{-1/2} \phi_0$ without dividing by the spatial width, since no particular extrapolation length is meant for the correction. Indeed, it will be possible to determine the actual extrapolation length only at convergence. The expression for the right boundary is straightforward, implying a non-negative current. Reflection is always reproduced at the center in curvilinear geometries.

About the currents from Eq. \ref{eq:J_tr}, the boundary conditions must be implemented through the first two terms at the right side due to the incoming flux. They are zero by definition only in case of vacuum. The flux expansion on the Legendre polynomials is necessary to reproduce other types of boundary condition. Again, we can only approximate the flux up to the first order, that is $\varphi \approx \phi/2 + 3/2 J \mu$. This approximation may not reproduce a vanishing current with reflection reproducing a symmetric distribution. At the left boundary for instance, the current can be expressed as $J(a) =  J^+ + J^- = 0$ with the entering current $J^+ = \int_0^1 { d\mu \mu \phi(a, \mu) } = - J^-$. $J^-$ takes into account all contributions coming from the right of $a$ up to $b$, see Eq. \ref{eq:J_tr}. In the absence of higher moments, the angular flux at the boundary becomes simply $\phi(a, \mu) \approx \phi_0(a) / 2$, yielding $J^+(x) = E_3 \left(\tau(a, x)\right)\phi_0(a)/2$. This approximation does not guarantee to obtain the expected vanishing current at the boundary when computing the same current at the boundary with symmetric flux distributions. However, it is possible to determine the difference between the partial currents (remind that $J^-$ is negative). The higher (even) flux moments are responsible for this residual quantity, but they are not available unfortunately. A possible solution is considering that this quantity comes artificially from the only second moment, like\footnote{The operator $\pluseq$ assigns to $J$ the sum of itself and quantity to the right.}
\begin{equation}
\begin{split}
J^+(x) &\pluseq \frac{5}{4} \tilde{\phi}_2(a) \left[ 3 E_5(\tau(a, x)) - E_3(\tau(a, x)) \right],\\ &\text{ with } \tilde{\phi}_2(a) = - \frac{16}{5} \left( \frac{1}{4} \phi_0(a) + J^- \right).
\end{split}
\end{equation}
The flux expansion at the left boundary is then
\[ \phi(a, \mu) = \frac{1}{2} \phi_0 + \frac{3}{2} \mu \phi_1 + \frac{5}{4} \left( 3 \mu^2 - 1 \right) \tilde{\phi}_2, \]
with $\phi_1 = J = 0$ with reflection. The derivation of the correction at the other boundary of the slab can easily be derived. Although this correction definitely improves the results, it cannot yield exactly the same results of the unfolded geometry, that is without reproducing the half symmetry by reflection. This occurs because the small current difference is allotted only to the second moment, neglecting the others which may become relevant on a case-dependent basis.

@@

The integral transport equation for the scalar flux can be written using the first flight collision probabilities as \cite{hebert2009applied}:
\begin{subequations}
\label{eq:cmp_eqn}
\begin{equation}
\label{eq:cmp1_eqn}
V_i \sigma_i \phi_i = \sum_{j=1}^I {q_j V_j \tilde{p}_{ij}}
                    + J^{(-)}_{b}S_{b} \tilde{p}_{ib},
\end{equation}
where $b$ stands for any boundary with a non-zero entering current. $b=I+1/2$ for the cylindrical and for the spherical geometries, whereas two incoming currents can be present in the slab. This equation can be simplified further by using the reciprocity property from Eqs. \ref{eq:reciprocity},
\begin{equation}
\label{eq:cmp2_eqn}
\phi_i = \sum_{j=1}^I {q_j p_{ji}} + J^{(-)}_{b} \tilde{e}_{bi}
\end{equation}
\end{subequations}
with the reduced escape probability $p_{ji} = \tilde{p}_{ji} / \sigma_j$ and the escape probability counting for both the signed terms. The incoming current at the boundary can be given as fixed external source, or being expressed as proportional to the out-going current by means of the albedo $0 \leq \beta \leq 1$, like
\begin{equation}
\label{eq:albedo}
  J^- = \beta J^+ = \beta \left(
      \sum_{j=1}^I {q_j p_{jb}} + \tilde{t} J^-
    \right),
\end{equation}
with the reduced collision probability from the surface at $x_b$ thanks again to reciprocity. About \ref{eq:reciprocity2}, reciprocity considers entering neutrons distributed according to the cosine described by the direction of flight and the normal to the same surface, yielding a cosine current by $J^\pm = \phi/4$, suitable to reproduce white reflection with $\beta = 1$. %yields the white reflection, that is isotropic inwards redistribution of the outgoing current, regardless of the outgoing direction.
The transmission $\tilde{t}$ is the probability that a neutron enters the outer surface with the cosine distribution, travels  inwards and leaves it back again without colliding. Conservation implies that $\tilde{t} = 1 - \sum_j{\tilde{p}_{jb}}$.

@@

In case of partial currents given by an albedo relation at the boundary, the same solving scheme still holds after redefinition of the matrix $\mathbf{P}$:
\[
\mathbf{P} \leftarrow \mathbf{P} + \frac{\beta}{1 - \beta \tilde{t}} \mathbf{E},
\]
with $\mathbf{E}=[\tilde{e}_{bi} p_{jb},\; 1 \leq i,j \leq I]$.

>>>>

%
% FIRST AUTHORS
%
\setlength{\baselineskip}{12pt}
\textbf{Daniele Tomatis}\\
DES, Service d'{\'e}tudes des r\'eacteurs et de math\'ematiques appliqu\'ees (SERMA),\\
CEA, Universit\'e Paris-Saclay, F-91191 Gif-sur-Yvette, France\\
daniele.tomatis@cea.fr\\
\vspace{0.25in}
\textbf{Roy Gross and Erez Gilad}\\
The Unit of Nuclear Engineering\\
Ben-Gurion University of the Negev, Israel\\
roygros@post.bgu.ac.il, gilade@bgu.ac.il\\
\end{center}

@@

Particle conservation on the full domain requires $\sum_i \tilde{p}_{ij} + \tilde{e}^+_{I+1/2,j} = 1$, $\forall j$. At last, reciprocity implies
\begin{subequations}
\label{eq:reciprocity}
\begin{gather}
  \frac{\tilde{p}_{ij}}{V_i \sigma_i} = \frac{\tilde{p}_{ji}}{V_j \sigma_j} \quad \text{and} \label{eq:reciprocity1}\\
  \frac{\tilde{p}_{j, i+1/2}}{V_j \sigma_j} = 4 \frac{\tilde{e}^+_{i+1/2, j} + \tilde{e}^-_{i+1/2, j}}{S_{i+1/2}},
  \label{eq:reciprocity2}
\end{gather}
\end{subequations}
for $1 \leq i,j, \leq I$. % and with the escape probability counting for both contributions from partial currents. These expressions
Eq. \ref{eq:reciprocity1} can also be used to verify the implementation of the escape probabilities. Collision probabilities with surface source emission from Eq. \ref{eq:reciprocity2} are needed with non-zero entering currents of neutrons at the boundary $x_{I+1/2}$. This probability refers again to an angular flux entering the surface $(i+1/2)$ isotropically, yielding the cosine current typical of white reflection.

@@

A generalized form for the boundary condition of the diffusion equation applying at the left side of the slab follows as $\Upsilon = - D_0 \phi_0 / (\Delta_0 / 2 + \zeta)$, where $\zeta$ is the extrapolation length in case of vacuum. $\zeta = 2D$ yields the Marshak boundary conditions for instance, but the value $\zeta = 2.13D$ is usually recommended in slab geometry \cite{stamm1983methods}. Reflection can be reproduced by $\zeta \rightarrow \infty$, whereas the condition of zero-flux comes with $\zeta = 0$. The boundary $\delta J$ takes the simpler form $\delta J = - \delta D_{-1/2} \phi_0$ without dividing by the spatial width, since no particular extrapolation length is meant for the correction. Indeed, it will be possible to determine the actual extrapolation length only at convergence. The expression for the right boundary is straightforward, implying a non-negative current. Reflection is always reproduced at the center in curvilinear geometries.

About the currents from Eq. \ref{eq:J_tr}, the boundary conditions must be implemented through the first two terms at the right side due to the incoming flux. They are zero by definition only in case of vacuum. The flux expansion on the Legendre polynomials is necessary to reproduce other types of boundary condition. Again, we can only approximate the flux up to the first order, that is $\varphi \approx \phi/2 + 3/2 J \mu$. This approximation may not reproduce a vanishing current with reflection reproducing a symmetric distribution. Only situations with perfectly isotropic angular flux at the boundary will be described correctly. At the left boundary for instance, the current can be expressed as $J(a) =  J^+ + J^- = 0$ with the entering current $J^+ = \int_0^1 { d\mu \mu \phi(a, \mu) } = - J^-$. $J^-$ takes into account all contributions coming from the right of $a$ up to $b$, see Eq. \ref{eq:J_tr}. In the absence of higher moments, the angular flux at the boundary becomes simply $\phi(a, \mu) \approx \phi_0(a) / 2$, yielding $J^+(x) = E_3 \left(\tau(a, x)\right)\phi_0(a)/2$, or $J^+(x) = \tilde{t}_a(x) \phi(a)/4$ using the CPM formalism (white reflection at $a$). Here, $\tilde{t}$ is the transmission probability of neutrons crossing uniformly and isotropically the surface $a$ to get to the position $x$ uncollided. The value of this probability can be obtained through reciprocity and conservation properties, as shown in section \ref{sec:cp}. Hence, this approximation does not guarantee to obtain the expected vanishing current at the boundary when computing the same current at the boundary with symmetric flux distributions. However, it is possible to determine the difference between the partial currents (remind that $J^-$ is negative). The higher (even) flux moments are responsible for this residual quantity, but they are not available unfortunately. A possible solution is considering that this quantity comes artificially from the only second moment, like\footnote{The operator $\pluseq$ assigns to $J$ the sum of itself and quantity to the right.}
\begin{equation}
\begin{split}
J^+(x) &\pluseq \frac{5}{4} \tilde{\phi}_2(a) \left[ 3 E_5(\tau(a, x)) - E_3(\tau(a, x)) \right],\\ &\text{ with } \tilde{\phi}_2(a) = - \frac{16}{5} \left( \frac{1}{4} \phi_0(a) + J^- \right).
\end{split}
\end{equation}
The flux expansion at the left boundary is then
\[ \phi(a, \mu) = \frac{1}{2} \phi_0 + \frac{3}{2} \mu \phi_1 + \frac{5}{4} \left( 3 \mu^2 - 1 \right) \tilde{\phi}_2, \]
with $\phi_1 = J = 0$ with reflection. The derivation of the correction at the other boundary of the slab can easily be derived.

Transmission probabilities of neutrons entering with anisotropic (even) distributions are now needed, but this time they can not be derived from the escape probabilities based on isotropic emission. We only give the expression to calculate these quantities in the slab, since white reflection is generally preferred in the 1-D curvilinear geometries for reproducing more accurate physical results.

Although this correction definitely improves the results, it cannot yield exactly the same results of the unfolded geometry, that is without reproducing the half symmetry by reflection. This occurs because the small current difference is allotted only to the second moment, neglecting the others which may become relevant on a case-dependent basis.

@@

The integral transport equation for the scalar flux with isotropic sources can be written using the first flight collision probabilities as \cite{lewis1984computational,hebert2009applied}:
\begin{subequations}
\label{eq:cmp_eqn}
\begin{equation}
\label{eq:cmp1_eqn}
V_i \sigma_i \phi_i = \sum_{j=1}^I {q_j V_j \tilde{p}_{ij}}
                    + J^{(-)}_{b}S_{b} \tilde{p}_{ib},
\end{equation}
where $b$ stands for any boundary with a non-zero entering current $J^{(-)}_{b} = \phi_b / 4$, given as well by an isotropic angular flux. $\tilde{p}_{ij}$ and $\tilde{p}_{ib}$ are the probabilities of a neutron to have its first collision in the volume $V_i$ after being emitted isotropically in the volume $V_j$ or from the surface $S_b$, respectively. $b=I+1/2$ for the cylindrical and for the spherical geometries after placing the center at $x_a = 0$ ($a=-1/2$), whereas two incoming currents can be present in the slab. This equation can be simplified further by using the reciprocity property from Eqs. \ref{eq:reciprocity},
\begin{equation}
\label{eq:cmp2_eqn}
\phi_i = \sum_{j=1}^I {q_j p_{ji}} + 4 J^{(-)}_{b} \tilde{e}_{bi}
\end{equation}
\end{subequations}
with the reduced escape probability $p_{ji} = \tilde{p}_{ji} / \sigma_j$ and the escape probability counting for both the signed terms. The incoming current at the boundary can be given as fixed external source, or being expressed as proportional to the out-going current by means of an albedo $0 \leq \beta \leq 1$, like
\begin{subequations}
\label{eq:albedo}
\begin{align}
\label{eq:albedo_b}
 % J^- = \beta J^+ = \beta \left(
 %     \sum_{j=1}^I {\frac{q_j p_{jb}}{4}} + \tilde{t} J^-
 %   \right),
  J_b^- &= \beta_b J_b^+ = \beta_b \left(
      \sum_{j=1}^I {\frac{q_j p_{jb}}{4}} + \tilde{r}_b J_b^- + \tilde{t}_{ba} \frac{S_a}{S_b} J_a^+
    \right), \text{ and}\\
\label{eq:albedo_a}
  J_a^+ &= \beta_a J_a^- = \beta_a \left(
      \sum_{j=1}^I {\frac{q_j p_{ja}}{4}} + \tilde{r}_a J_a^+ + \tilde{t}_{ab} \frac{S_b}{S_a} J_b^-
    \right),
\end{align}
\end{subequations}
with the reduced collision probability from the surface at $x_b$ thanks again to reciprocity. About \ref{eq:reciprocity2}, reciprocity always considers entering neutrons distributed according to the cosine described by the direction of flight and the normal to the same surface, yielding a cosine current by $J^\pm = \phi/4$, suitable to reproduce white reflection with $\beta = 1$. %yields the white reflection, that is isotropic inwards redistribution of the outgoing current, regardless of the outgoing direction.
The reflection $\tilde{r}$ is the probability that a neutron enters the outer surface with the cosine distribution, travels inwards and leaves it back again without colliding. It must be zero in the slab, but not in the cylinder or in the sphere. Transmission is used instead when the leaving surface to be crossed is different from the entering one, and it is non-vanishing here only in the slab. Conservation applied in the cylinder or in the sphere implies that $\tilde{r}_b = 1 - \sum_j{\tilde{p}_{jb}}$, where we have also assumed implicitly no possible transmission in the derivation of their escape probabilities. Similarly in the slab, it is $\tilde{t}_{ab} = 1 - \sum_j{\tilde{p}_{jb}}$ for instance. Eq. \ref{eq:albedo_a} can be used for Eq. \ref{eq:albedo_b} to obtain an expression for $J_b^-$ that is suitable for substitution in Eq. \ref{eq:cmp1_eqn}. Eqs. \ref{eq:albedo} are in a general form for 1-D geometries, which simplify either in the slab or in the others.

@@

In case the partial current is given by an albedo relation at the boundary of the cylinder (or of the sphere), the same solving scheme still holds after redefinition of the matrix $\mathbf{P}$:
\[
\mathbf{P} \leftarrow \mathbf{P} + \frac{\beta_b}{1 - \beta_b \tilde{r}} \mathbf{E}_b,
\]
with $\mathbf{E}_b = [\tilde{e}_{bi} p_{jb},\; 1 \leq i,j \leq I]$. This redefinition holds also for the slab in case of periodic boundary condition, i.e. $J_b^- = J_a^+$. Lastly for the slab, a general relation for the right boundary is
%\[
%J_b^- = \beta_b \left[
%      \sum_{j=1}^I {\frac{q_j p_{jb}}{4}} + \tilde{t} \beta_a \left(
%      \sum_{j=1}^I {\frac{q_j p_{ja}}{4}} + \tilde{t} J_b^- \right)
%    \right]
%\]
\[ \mathbf{P} \leftarrow \mathbf{P} + 
\frac{\beta_b}{1 - \beta_a \beta_b \tilde{t}^2}
\left( \mathbf{E}_b + \beta_a \tilde{t} \mathbf{E}_a \right)
\]
with $\mathbf{E}_a = [\tilde{e}_{bi} p_{ja},\; 1 \leq i,j \leq I]$. The corresponding expression at the left boundary is easy to obtain.
